Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-21 docs/proposals/NSF/ProjDesc/ProjDesc.tex
checked off Nick's comments
NM

{\bf This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow alters porous-media properties and to characterize the associated changes in dispersive transport within the evolving medium.} A recognized complexity of naturally occurring media is the heterogeneity and anisotropy of macroscopic properties such as permeability and mechanical strength~\cite{neuman1987stochastic, moyner2016multiscale, lin2018randomization}. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.  Physical mechanisms that can alter porous-medium properties include: 
\begin{itemize}[noitemsep,topsep=0pt]
\item Erosion of solid material due to fluid-mechanical stresses or chemical dissolution.
\item Transport of grains due to intervening fluid flow and due to sedimentation.
\item Compaction of the medium due to overlying loads.
\item Gravitational collapse of a medium due to loads exceeding a threshold.
\end{itemize}
The above are roughly ordered according to timescale, from the slowest acting to the fastest. While many of these mechanisms have been studied individually in various contexts and to varying levels of rigor, very little is known about their combined effect on the dynamic evolution of a porous medium. Since we aim to build a quantitative model from the ground up, we will first focus on the slowest acting mechanisms, as these ultimately set the stage for the more rapid events like sinkhole collapse.

In particular, we view the slow, accumulated effects of erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of porous-medium constituents, enabling smaller fragments to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of fragments or grains. For instance, erosion can carve grains  into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties~\cite{szymczak2009wormhole}. Some visible effects of flow-induced erosion are shown in Fig.~\ref{collage}. The top row shows products of primarily {\em chemically-driven} erosion, i.e.~dissolution, while the bottom shows effects of primarily {\em mechanically-driven} erosion. Some examples in the figure include the rough surface of limestone pavement, a sinkhole relic in Belize, and an erosion-curved conduit. The final image shows the small-scale structure of soil which exhibits polydispersity and anisotropy resulting from a combination of erosive, compaction, and transport actions.

@@

\item The Stokes limit introduces {\em longer} range effects than the inertial limit does, thus coupling the shape evolution of {\em all} grains and conduits to a higher degree.

@@

\todo[inline]{Okay, I did some rearranging. See if you like it better.}

@@

We first outline the theoretical foundation that will be used in later sections of the proposal. At the finest scale, the fluid flow intervening between individual grains of a porous medium is governed by the incompressible Stokes equations
\begin{align}
  \label{eqn:StokesEq}
  \mu \Delta \uu = \grad p, \quad
  \grad \cdot \uu = 0, \quad \xx \in \Omega,
\end{align}
%\begin{equation}
%\label{eqn:StokesEq}
%\begin{split}
%\mu \Delta \uu = \grad p,	&\hspace{20pt} \xx \in \Omega, \\
%\grad \cdot \uu = 0, 		&\hspace{20pt} \xx \in \Omega,
%\end{split}
%\end{equation}
where $\uu$ is the velocity field, $p$ the pressure field, $\mu$ the fluid viscosity, and $\Omega$ the fluid domain. Linear elliptic PDEs, such as the Stokes equations, can be reformulated as a boundary integral equation (BIE).  BIEs have several numerical advantages including a dimension reduction, automatic mass conservation, and high-order accuracy. An equivalent BIE formulation of the Stokes equations~\eqref{eqn:StokesEq} is
\begin{align}
  \label{eqn:BIE}
  \ff(\xx) = -\frac{1}{2}\eeta(\xx) + \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy, \quad \xx \in \bd\Omega,
\end{align}
where $\ff$ is a Dirichlet boundary condition, $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the outward unit normal of $\bd\Omega$.  Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, quantities such as the velocity, pressure, vorticity, and deformation tensor can be computed in the fluid bulk.  For example, the velocity at $\xx \in \Omega$ is 
\begin{align}
  \uu(\xx) = \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy.
\end{align}

@@

While the tortuosity characterizes the amount of winding of trajectories, the dispersion characterizes their spreading.  Dispersion can be described using the first- and second-ensemble moments of the trajectories
\begin{align}
  \langle \lambda \rangle (t) = \frac{1}{N_p} 
      \sum_{j=1}^{N_p} \lambda_j(t), \quad
  \sigma_\lambda^{2}(t) = \frac{1}{N_p}
    \sum_{j=1}^{N_p} (\lambda_j(t) - \langle \lambda \rangle(t))^2.
\end{align}
\begin{wrapfigure}[22]{l}{0.4\textwidth}
  \includegraphics[width=0.4\textwidth]{figs/100b_second_moment_long_ref}
  \caption{\label{fig:100dispersion} \em The particle spreading, characterized with the standard deviation $\sigma_\lambda(t)$ at six different porosities. At early times, the line of slope 1 (black dashed curve) indicates a ballistic motion. At later times, the dispersion becomes super-dispersive at a rate indicated by the slope of the black dashed-dotted line.}
\end{wrapfigure}
Then, the standard deviation $\sigma_\lambda$ characterizes the dispersion, and long-time behavior typically results in a power-law scaling $\sigma_\lambda \sim t^{\alpha}$, where $\alpha$ quantifies the long-time spreading. At early times, transport is ballistic with $\alpha = 1$, as the trajectories have not explored much of the porous geometry.  Once the trajectories have traversed a few grains, however, spreading typically becomes super-dispersive with $\alpha \in (0.5,1)$~\cite{dea-leb-den-tar-bol-dav2013}. In Fig.~\ref{fig:100dispersion}, we investigate spreading at several different porosities of a porous geometry that is initialized with 100 grains.

>>>>

{\bf This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow alters porous-media properties and to characterize the associated changes in dispersive transport within the evolving medium.} A recognized complexity of naturally occurring media is the heterogeneity and anisotropy of macroscopic properties such as permeability and mechanical strength~\cite{neuman1987stochastic, moyner2016multiscale, lin2018randomization}. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.  Physical mechanisms that can alter porous-medium properties include: 
\begin{itemize}[noitemsep,topsep=0pt]
\item Erosion of solid material due to fluid-mechanical stresses or chemical dissolution.
\item Transport of fragments and grains due to intervening fluid flow and due to sedimentation.
\item Compaction of the medium due to overlying loads.
\item Gravitational collapse of a medium due to loads exceeding a threshold.
\end{itemize}
The above are roughly ordered according to timescale, from the slowest acting to the fastest. While many of these mechanisms have been studied individually in various contexts and to varying levels of rigor, very little is known about their combined effect on the dynamic evolution of a porous medium. Since we aim to build a quantitative model from the ground up, we will first focus on the slowest acting mechanisms, as these ultimately set the stage for the more rapid events like sinkhole collapse.

In particular, we view the slow, accumulated effects of erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of porous-medium constituents, enabling smaller fragments to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of individual fragments or particles. For instance, erosion can carve particles into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties~\cite{szymczak2009wormhole}. Some visible effects of flow-induced erosion are shown in Fig.~\ref{collage}. The top row shows products of primarily {\em chemically-driven} erosion, i.e.~dissolution, while the bottom shows effects of primarily {\em mechanically-driven} erosion. Some examples in the figure include the rough surface of limestone pavement, a sinkhole relic in Belize, and an erosion-curved conduit. The final image shows the small-scale structure of soil which exhibits polydispersity and anisotropy resulting from a combination of erosive, compaction, and transport actions.

@@

\item The Stokes limit introduces {\em longer} range effects than the inertial limit does, thus coupling the shape evolution of {\em all} fragments, grains, and conduits to a higher degree.

@@

@@

We first outline the theoretical foundation that will be used in later sections of the proposal. At the finest scale, the fluid flow intervening between individual constituents of a porous medium is governed by the incompressible Stokes equations
\begin{align}
  \label{eqn:StokesEq}
  \mu \Delta \uu = \grad p, \quad
  \grad \cdot \uu = 0, \quad \xx \in \Omega,
\end{align}
%\begin{equation}
%\label{eqn:StokesEq}
%\begin{split}
%\mu \Delta \uu = \grad p,	&\hspace{20pt} \xx \in \Omega, \\
%\grad \cdot \uu = 0, 		&\hspace{20pt} \xx \in \Omega,
%\end{split}
%\end{equation}
where $\uu$ is the velocity field, $p$ the pressure field, $\mu$ the fluid viscosity, and $\Omega$ the fluid domain. Linear elliptic PDEs, such as the Stokes equations, can be reformulated as a boundary integral equation (BIE).  BIEs have several numerical advantages including a dimension reduction, automatic mass conservation, and high-order accuracy. An equivalent BIE formulation of the Stokes equations~\eqref{eqn:StokesEq} is
\begin{align}
  \label{eqn:BIE}
  \ff(\xx) = -\frac{1}{2}\eeta(\xx) + \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy, \quad \xx \in \bd\Omega,
\end{align}
where $\ff$ is a Dirichlet boundary condition, $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the outward unit normal of $\bd\Omega$.  Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, quantities such as the velocity, pressure, vorticity, and deformation tensor can be computed in the fluid bulk.  For example, the velocity at $\xx \in \Omega$ is 
\begin{align}
  \uu(\xx) = \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy.
\end{align}

@@

While the tortuosity characterizes the amount of winding of trajectories, the dispersion characterizes their spreading.  Dispersion can be described using the first- and second-ensemble moments of the trajectories
\begin{align}
  \langle \lambda \rangle (t) = \frac{1}{N_p} 
      \sum_{j=1}^{N_p} \lambda_j(t), \quad
  \sigma_\lambda^{2}(t) = \frac{1}{N_p}
    \sum_{j=1}^{N_p} (\lambda_j(t) - \langle \lambda \rangle(t))^2.
\end{align}
\begin{wrapfigure}[23]{l}{0.4\textwidth}
  \includegraphics[width=0.4\textwidth]{figs/100b_second_moment_long_ref}
  \caption{\label{fig:100dispersion} \em The particle spreading, characterized with the standard deviation $\sigma_\lambda(t)$ at six different porosities. At early times, the line of slope 1 (black dashed curve) indicates a ballistic motion. At later times, the dispersion becomes super-dispersive at a rate indicated by the slope of the black dashed-dotted line.}
\end{wrapfigure}
Then, the standard deviation $\sigma_\lambda$ characterizes the dispersion, and long-time behavior typically results in a power-law scaling $\sigma_\lambda \sim t^{\alpha}$, where $\alpha$ quantifies the long-time spreading. At early times, transport is ballistic with $\alpha = 1$, as the trajectories have not explored much of the porous geometry.  Once the trajectories have traversed a few grains, however, spreading typically becomes super-dispersive with $\alpha \in (0.5,1)$~\cite{dea-leb-den-tar-bol-dav2013}. In Fig.~\ref{fig:100dispersion}, we investigate spreading at several different porosities of a porous geometry that is initialized with 100 grains.
